---
title: Principles of Data Munging & Wrangling
author: Jan
tags: data munging, dplyr
---

Terms like data *munging* or *wrangling* have come to refer to the act of preparing raw data for further processing.
This often involves transformation into a format that is supported by the analysis software of choice, discarding unnecessary information, re-arranging the layout of the data and replacing/reformatting entries to facilitate processing.
Deriving new measures from one or several entries could already count as 'analysis' but is often seen as part of pre-processing if simple enough.
In the following I'll describe each of the <em>munging/wrangling</em> categories and illustrate them with a practical R example.

<div></div><!--more-->

We'll follow this [dplyr](http://seananderson.ca/2014/09/13/dplyr-intro.html) introduction closely and incorporate additional sources[^konduforov].
Let's go through each of the ten or so major <em>munging/wrangling</em> steps.

* import/conversion
* exploration
* renaming
* subsetting/filtering
* ordering/sorting
* rearranging
* mutating/summarizing


There is no strict order to these actions.
It is even quite common to repeat and alternate between these steps until the data is in 'shape'.

[^konduforov]: [Fast data munging in R](http://www.slideshare.net/AlexMerle/data-munging-r).


## Import and Conversion

Data *import* refers to reading data into memory, usually from a file.
Let's start by downloading a data set and reading it into [R](https://www.r-project.org).

```{r eval=FALSE}
url <- 'http://esapubs.org/archive/ecol/E090/184/PanTHERIA_1-0_WR05_Aug2008.txt'
t <- read.table(url, header=T, sep='\t', check.names=F)
```

It might be more convenient to download the 2.2 MB large data set to a local directory using `curl -O URL` on OS X or `wget URL` on Linux and then run the following instead.


```{r eval=TRUE}
url <- '/tmp/PanTHERIA_1-0_WR05_Aug2008.txt'
t <- read.table(url, header=T, sep='\t', check.names=F)
```

Although it seems as if we've only copied or imported the data from a remote source, the data has already been interpreted and converted to an internal representation.
This happened magically with the `read.table` command, which reads or 'scans' a text file of whitespace-separated columns into a `R` data frame.
Type `?read.table` on the `R`-prompt for an explanation of the arguments.
In a nutshell: `header=T` indicates that the first row should be used for the column names, `sep=\t` indicates that the columns are separated by tabulators (but not spaces), and `check.names=F` prevents `R` from replacing characters such as spaces with periods.
These options could be seen as 'hints' that help a parser to interpret the information.

It is surprisingly common to come across data formatted as a table in a text file, because these files can be easily viewed and understood (just open them in a text editor).
It is also very easy to write these files out or parse them using only a few lines of code in most programming languages.
In contrast, binary encoding or markup languages allow to store data that does not conform to 2D-tables.
However, these formats usually require more complicated 'readers' and 'writers'.
This can lead to *format hell*[^formathell], especially with proprietary formats.

[^formathell]: The problem where format A needs to be converted to format D, but available converters only support conversion from A to B or C to D. The user is than forced to find a converter B to D or A to C, which is equivalent to the problem the user started out with.


## Exploration

Before we continue we should quickly to check that the data has been successfully imported into `R`.
Let's print the first five rows of the first three columns:

```{r comment=''}
t[1:5,5:7]
```

Apparently we have different types of data, such as *strings*, *integers*, and *floats* (i.e. floating point numbers).
Altogether we have `r nrow(t)` rows and `r ncol(t)` columns.

```{r comment=''}
dim(t)
```

We can explore the data set by printing individual columns and using the `head`, `tail` `summary`, and `str` functions.
This should quickly reveal that the value `-999` seems to have some special meaning, possibly indicating a missing entry.
`R` supports missing entries with the `NA` value.

Let's pretend we are back in the import/conversion section and set `-999` to `NA` throughout.

```{r comment=''}
t[t==-999] <- NA
```

As we explore the data set we might make further adjustments related to import/conversion, such as changing the *type* of a column (by using `as.numeric`, `as.character`, `as.factor`, etc.).
Sometimes it is also desirable to substitute characters (by using `gsub`).
For example, column names that contain spaces or other special characters can sometimes complicate processing further down the line.

Sometimes a plot is worth a thousand characters.
It is helpful to `plot` columns or their histograms (`hist`).
Let's plot a histogram of the body mass column.


```{r comment='', fig.width=6, fig.height=4, fig.cap=""}
hist(log(t$`5-1_AdultBodyMass_g`), main='log adult body mass [g]', xlab='')
```


## Subsetting/Filtering

We are often only interested in a subset of the data.
As long as we are not running out of memory there is no technical reason to remove data.
However, it usually makes life easier if we remove 'clutter' that obscures relevant information when a data frame is printed on the screen.
In addition, removing irrelevant information in advance can simplify or remove the need for indexing or subsetting expressions later on.

We'll be using the [dplyr](https://github.com/hadley/dplyr) package in the following.

```{r comment='', message=FALSE}
library(dplyr)
t <- dplyr::tbl_df(t) # pretty print
```

Using `select` and `filter` functions we can keep or remove specific columns and rows.


### Select
Columns can be selected by matching their names with `starts_with`, `ends_with`, `contains`, `matches `.
Prefixing a `-` will drop the matched column(s).

```{r comment=''}
# select all columns with 'body' in their name
select(t, contains("body"))[,1:3]

# drop these columns
select(t, -contains("body"))[,1:3]
```


### Filter
The `filter` function filters rows based on values, often involving logical expressions >, <, ==, &, |, etc.

```{r comment=''}
# select rows with large BMI
filter(t, `5-1_AdultBodyMass_g` > 1e6)[,c(1:3,7)]

# restrict to a specific 'order'
filter(t, `MSW05_Order`=='Carnivora' & `5-1_AdultBodyMass_g` > 1e6)[,c(1:3,7)]
```



## Ordering/Sorting
Ordering or sorting a data frame is usually not necessary for the actual analysis, it helps, however, with planning and confirming filtering operations.


```{r comment=''}
arrange(t, `MSW05_Order`)[,c(1:3,7)]
# arrange(t, desc(`MSW05_Order`))[,c(1:3,7)]
```


## Rearranging

melt, cast


## Renaming

No text.


## Mutating/Summarizing

### Mutating
Adding values derived (usually in a simple manner) form entries in the data to facilitate further analysis.

```{r comment=''}
#glimpse(mutate(t, g_per_mm = adult_body_mass_g / adult_head_body_len_mm))
```


### Summarizing
Creating a new table that collapses information into summaries according to grouping factors.
For example, one can calculate the mean body mass for each `order`.


```{r comment=''}
summarise(group_by(t, `MSW05_Order`), 
          mean_mass = mean(`5-1_AdultBodyMass_g`, na.rm = TRUE))
```


## Piping Data

The pipe operator `%>%` allows to connect several of the above functions.
It is essentially a better way to write hard to parse *nested braces* expressions $h(g(f(x)))$ in a more natural way that mimics the *flow* of the date: $x \rightarrow f \rightarrow g \rightarrow h$

Let's recapitulated [Sean Anderson's](http://seananderson.ca/2014/09/13/dplyr-intro.html) example and find the species with the highest body-mass-to-length ratio:


```{r comment=''}
t %>%
 mutate(mass_to_length=`5-1_AdultBodyMass_g`/`13-1_AdultHeadBodyLen_mm`) %>%
 arrange(desc(mass_to_length)) %>%
 select(`MSW05_Species`, mass_to_length)
```

Although this is a very simple example it already illustrates how the different *munging* and *wrangling* lego bricks can be stacked together to create the data frame(s) necessary for the model fitting and further analyses.



